---
title: "Lab Report Template (Version 1)"
author: "Your Name Here"
format:
  html:
    df_print: "paged"
    fig_caption: yes
    toc: true #include a table of contents (optional)
    toc_depth: 3
    number-sections: true
    self-contained: true #save images etc. in this file (vs folders)
execute:
  echo: false # hide raw code output
  cache: false
  echo.comments: false
  message: false
  warning: false

---

`r format(Sys.time(), '%B %d, %Y')`.

---

# Notes on This Document

* This template is designed as an example for structuring your lab reports. It provides a resource to support your work but is **entirely optional and not the only acceptable format for lab report submissions**. Developing your scientific judgment is a key goal of this class. 
  * For example, you may choose to separate the Results and Discussion into distinct sections if that better suits your analysis and writing style.   
* Please refer to the 'Lab Report Guidelines.html' document for detailed requirements on structuring and formatting each section.    
* Make sure to add comments to your code to aid in transparency and future replication   
* Before using this template, delete: 
  * this Notes on This Document Section   
  * things in brackets 
  * irrelevant text
* This document provides suggested paragraph lengths for each section. Academic paragraphs are roughly 100-300 words, but use your judgment. 

---

```{r}
# ----- clear work space of all objects and unload all extra (non-base) packages --------------
rm(list = ls(all = TRUE))
if (!is.null(sessionInfo()$otherPkgs)) {
    res <- suppressWarnings(
        lapply(paste('package:', names(sessionInfo()$otherPkgs), sep=""),
               detach, character.only=TRUE, unload=TRUE, force=TRUE))
   
}


# Load pacman into memory, installing as needed
my_repo <- 'http://cran.r-project.org'
if (!require("pacman")) {install.packages("pacman", repos = my_repo)}

# ----- load packages with pacman -----
# Load additional the required packages, as needed
pacman::p_load(tidyverse)

# ----- load in any required scripts ----- 
#source("global_functions.R")

```

```{r}
# ----- code to load data -----
# Example: my_data <- read.csv("data/my_data_file.csv")
# Make sure to specify the correct file path

```

```{r}
# ----- clean data -----


```


# Introduction        
[~2-4 paragraphs]

[Provide background information, a clear statement of objectives, and any relevant hypotheses.]



In this lab, we aim to....


# Methods   
[1-3 paragraphs total in the following subsections]

[Describe the steps you took to conduct the analysis, providing enough detail for a knowledgeable expert to replicate your approach.]

## Data Description   
[~1 paragraph]
[Summarize the data used in these analyses.]


## Statistical Analyses    
[~1 paragraph]


# Results & Discussion  
[~1-3 paragraphs per question]    

## Question 1 [Or descriptive header]

[Explain findings and any trends observed in the results. Describe each result in the context of your hypotheses and/or objectives.]


We find that......



Table 1 shows ......

```{r}
# ----- code to generate table 1 -----

```

We also find that....., as shown in Figure 1.

```{r}
# ----- code to generate figure 1 -----



```


## Question 2 [Or descriptive header]

Moreover, in analyses evaluating.....we see that.....

```{r}
# ----- additional analyses ----- 




```



## Question 3 etc. 


```{r}
# ----- analyses for question 3 -----




```



# Conclusions    
[~1 paragraph]

In conclusion...



## Strengths & Limitations

[As relevant]


---

# Appendix

## Supplementary Information   
[Only include additional results or code that directly supports your findings. Each item should have a descriptive label and be referenced in the main text.]



## Session Information 

```{r session.info}
# ----- session information ----

# print R session information
sessionInfo()

```

## Analytic Code 

```{r code.appendix, ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE, , include=TRUE}
# ----- code appendix ----
```

## Functions

```{r functions, eval = TRUE}
# ----- functions ----

# Show the names of all functions defined in the .Rmd
# (e.g. loaded in the environment)
lsf.str()

# Show the definitions of all functions loaded into the current environment  
lsf.str() %>% set_names() %>% map(get, .GlobalEnv)

```

