---
title: "Week 5 Lab: Variance Components"
author: "Instructors for ENVH 556 Autumn 2024"
format:
  html:
    df_print: "paged"
    fig_caption: yes
    toc: true
    toc_depth: 3
    number_sections: true
    self-contained: true #save images etc. in this file 
editor_options: 
  chunk_output_type: console
---

This document was rendered on `r format(Sys.time(), '%B %d, %Y')`.

```{r setup, include=FALSE}
#-----setup-----
# clear work space of all objects and unload all extra (non-base) packages
rm(list = ls(all = TRUE))
if (!is.null(sessionInfo()$otherPkgs)) {
    res <- suppressWarnings(
        lapply(paste('package:', names(sessionInfo()$otherPkgs), sep=""),
               detach, character.only=TRUE, unload=TRUE, force=TRUE))
}

```

```{r load.libraries.pacman, echo=FALSE, include=FALSE, eval=TRUE}
#-----load libraries pacman-----

# Load pacman into memory, installing as needed
my_repo <- 'http://cran.r-project.org'
if (!require("pacman")) {install.packages("pacman", repos = my_repo)}

# NOTE: if you receive an error with e.g., fitVCA(), you may need to reinstall lme4
# install.packages("lme4", repos = "http://cran.r-project.org", type = "source")

# Load the other packages, installing as needed.
pacman::p_load(knitr, dplyr, tidyr, modelr, stringr, purrr, ggplot2, Hmisc, 
               lme4, VCA)

```

```{r directory.organization.read.data, echo=FALSE, warning=FALSE}
#-----directory organization and read data-----

# create "Datasets" directory if one does not already exist    
dir.create(file.path("Datasets"), showWarnings=FALSE, recursive = TRUE)

# specify data path
data_path <- file.path("Datasets")

# specify the file name and path
file_name <- "weldschool.rds"
file_path <- file.path(data_path, file_name)

# Download the file if it is not already present
if (!file.exists(file_path)) {
    url <- paste("https://faculty.washington.edu/sheppard/envh556/Datasets", 
                 file_name, sep = '/')
    download.file(url = url, destfile = file_path)
}

# Output a warning message if the file cannot be found
if (file.exists(file_path)) {
    weldschool <- readRDS(file_path)
} else warning(paste("Can't find", file_name, "!"))

# remove temporary variables
rm(url, file_name, file_path, data_path)

```

# Introduction and Purpose

The purpose of this lab is to **practice estimating variance components both by using descriptive statistics as well as through application of random and mixed effect models**. In addition, as with all the labs in this course, you should focus on writing up a coherent lab report that addresses the scientific context and scientific questions that can be answered using variance components.

In this lab exercise, we will use the Welding School personal survey data from the *Baker et al. 2014 Blood Manganese as an Exposure Biomarker Review*. The data are stored in the file named “**weldschool.rds**”. The exposure concentration variables in the dataset consist of:    

* *mnconc*: airborne concentrations of manganese (ug/m^3^)  
* *particulate*: airborne concentrations of total particulate (µg/m^3^)
* *subjid*: subject ID is the grouping variable 

Further description of the variables can be found in the document titled "*Welding School Exposures variables 556 rev021015*."

# Getting Started

The companion file *Week5Lecture.Rmd* gives R commands relevant to this lab. The lecture examples are applied to the DEMS underground and Snapshot traffic gradient site data. Unlike previous labs, we do not provide you with worked examples from the Welding School dataset. You will need to apply the tools demonstrated in the *Week5Lecture.Rmd* to your analyses for this lab.

## Tools for estimating variance components

Below is a summary of R packages. See the *Week5Lecture.Rmd* file for coding examples.

**There are different ways of estimating variance components.**

* Descriptively estimate variance components using `dplyr`
  * simple, helps understand the approach  intuitively   
* Random effects models only   
    - ANOVA method of moments using `VCA`
    - ANOVA restricted maximum likelihood using `VCA`
* Random effects and mixed models
    - ANOVA restricted maximum likelihood using `lmer` in the `lme4` package
      * What is often used in practice     
    - ANOVA maximum likelihood using `lmer` in the `lme4` package

## Overview of `lmer` commands

### `lmer` overview

**The `lmer` function in the `lme4` package looks similar to a standard linear regression model but adds capacity to incorporate structured errors as random effects**. You can fit a variety of random effects models, from the simplest (a single random intercept) to more complex models with multiple random effects terms, nested (vs. crossed) terms, and random slope models.

The formula part of the `lmer` command allows random effects to be specified with terms inside parentheses. **The vertical bar separates the formula part of the random effect model specification (before "|") from the variable that defines the random effect (after "|")**.

Fitting of `lmer` models is done by maximum likelihood estimation. The random effects and residual (unstructured) error in this model are assumed to have normal distributions. **The default estimation uses restricted maximum likelihood, which corrects for the bias in the variance component estimates**. However, **when comparing models with different numbers of terms, standard maximum likelihood should be used**.

### `lmer` random effects coding

Here is a [link](https://stats.stackexchange.com/questions/13166/rs-lmer-cheat-sheet) to a helpful stackexchange post on `lme4` random effects coding. Here is a brief summary for categorical (grouping) variable G and continuous variable X, where our goal is to have random effects for levels of G:

* Random intercept only: (1|G)
* Random slope only:  (0 + X|G)
* Correlated random slopes and intercepts: (1 + X|G)
* Independent random slopes and intercepts: (1|G) + (0 + X|G)

*Side Note: Longitudinal models with random slopes and intercepts often allow for correlated slopes and intercepts. In this situation, there will be a third parameter estimated to reflect that. We can force these to be independent, however, in the code.*

# Practice Session

1. Determine your project for this lab.

2. Read in the *weldschool.rds* dataset:
    a. **Become familiar with your data.**
    b. How many observations, students, and quarters are in your dataset? How many observations are there per student (average, minimum, maximum)?
    c. Characterize the outcome variables *mnconc* and *particulate*, airborne measurements of Mn concentration and total particulate, respectively, and all potential covariates.
    d. Get a basic understanding of the distribution of the airborne measurements on both the native and log-transformed scale. What scale do you think you should conduct your variance components analysis on?
    
For the rest of the practice session, we will focus on the Mn concentration variable:

3. Try to estimate the variance components within and between subjects "by hand" **using `dplyr`**.

4. Now **use `VCA`** to estimate the variance components using a method of moments or ANOVA approach.
    a. Do your estimates agree with those from Step 3? Why or why not?

5. Now **use `lmer`** to estimate the variance components. (Decide whether you want maximum likelihood or restricted maximum likelihood estimates. Be able to articulate your reason for your choice.)
    a. Do your estimates agree with those from Step 3? Why or why not?

6. Decide whether you should **add any fixed effects** into your model.
    a. If so, which one(s)? Why? (Note: Make sure to consider science in your answer!)
    b. **Compare** the variance components from the models with and without fixed effect(s). What happens to the variance component estimates from Step 5 after you add your selected covariate(s)? Does this make sense to you? (Hint: Consider how your covariate is distributed within and between subjects in the data.)

# Homework Exercises

1. **Repeat the practice session analysis with particulate concentration**. Does this produce any different results? Why or why not?

2. How do your variance component estimates compare with the DEMS and Snapshot examples we discussed in lecture?

3. Report and discuss the results of your analysis of Mn and total particulate for this lab, touching on the following points:
    a. The design of the study and structure of the dataset.
    b. The scale of the data on which you did your analysis.
    c. The comparability of variance component estimates using different estimation approaches and some possible reasons for (lack of) comparability.
    d. The impact on the variance component estimates when you include fixed effect(s) in your model and the reason for the change(s) in these data.
    e. Incorporate insights from the [Peretz et al, 2002](https://doi.org/10.1093/annhyg/mef009) paper.


# Appendix

```{r session.info}
#-----session information----

# print R session information
sessionInfo()

```

```{r code.appendix, ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE, , include=TRUE}
#-----code appendix----
```

```{r functions, eval = TRUE}
#-----functions----

# Show the names of all functions defined in the .Rmd
# (e.g. loaded in the environment)
lsf.str()

# Show the definitions of all functions loaded into the current environment  
lsf.str() %>% set_names() %>% map(get, .GlobalEnv)

```
